A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-16 tex/codatadp.tex
more

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables.
For example, 
%$\mu x. \const{C}( x ) \aequiv \mu y. \const{C}( y )$ and
$\mu x.\; \const{D}( y, x ) \aequiv \mu z.\; \const{D}( y, z )$,
but
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{D}( y, x ) \not\aequiv \mu x.\; \const{D}( z, x )
\not\aequiv \mu y.\; \const{D}( y, x )$.
Two $\mu$-terms may represent the same value while not being $\alpha$-equivalent, as in
$\mu x. \const{C}( \const{C}( x ) ) \not\aequiv \mu x. \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a co-datatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when $\tau \not\in \Codata$.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:

>>>>

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables.
For example, 
%$\mu x. \const{C}( x ) \aequiv \mu y. \const{C}( y )$ and
$\mu x.\; \const{D}( y, x ) \aequiv \mu z.\; \const{D}( y, z )$,
but
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{D}( y, x ) \not\aequiv \mu x.\; \const{D}( z, x )
\not\aequiv \mu y.\; \const{D}( y, x )$.
Two $\mu$-terms may represent the same value while not being $\alpha$-equivalent, as in
$\mu x.\; \const{C}( \const{C}( x ) ) \not\aequiv \mu x.\; \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a co-datatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when
%$\tau \not\in \Codata$.
$\tau \in \Data$.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:
